|

### 2021-08-11 prova_knit_finale.Rmd
Model comparison: changes and move to end Other changes

```{r eval=FALSE, include=FALSE}
truex<-x
estimatedx<-matrix(NA,ncol=5,nrow=5)
colnames(estimatedx)<-c("N","BPF","GPF", "APF","LWF")
estimatedx[,1]<-c("20","100","1000","10000","100000")

@@

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, results="asis"}
kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=4)
```

@@

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Riferimento alla sezione di convergenza?.} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500. Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}

@@

\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}\label{Forecast comparison}

In this section, we apply the particle filter in a model-comparison
exercise, ultimately underlining the importance of taking stochastic
volatility into account in modeling equity returns. We follow XX, comparing density forecasts through relative predictive and cumulative likelihoods.

We analyze a dataset of log close-to-close returns of the S&P500, from 2017 to 2021. Consider a stochastic volatility model
$$
\mathcal{M}_{SV}:\begin{array}{lc}
r_{t}=\alpha_{r}+\sigma_{r,t}\cdot v_{t}\\
\log\sigma_{r,t}^{2}=\alpha_{\sigma}+\beta_{\sigma}\cdot\log\sigma_{r,t-1}^{2}+\sigma_{\sigma}\cdot w_{t}
\end{array}\ (v_{t},w_{t})\sim\mathcal{N}_{2}(\boldsymbol{0},I_{2})
$$

Whereas the emission distribution $p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})$
is known, we approximate the predictive state distribution $p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})$
with weighted particles obtained with particle filtering.

Figure XX reports the time series of the cumulative log-likelihood
given the constant volatility model $\mathcal{M}_{CV}$ relative to
the one given $\mathcal{M}_{SV}$, obtained as $\log p(y_{1:t}|\mathcal{M}_{CV})-\log p(y_{1:t}|\mathcal{M}_{SV})$.
Notice that the first difference of this series is the relative log-likelihood
of the incoming observation, $p(y_{t}|y_{1:t-1},\mathcal{M}_{CV})-p(y_{t}|y_{1:t-1},\mathcal{M}_{SV})$.

```{r cll_plot, echo=FALSE, fig.align='center', fig.cap="Relative log-likelihood", fig.height=3, fig.pos='H', fig.width=10, results=F}
load("SV/cll_plot.Rda")
cll_plot+
  theme_bw()+
  scale_x_date(date_breaks = "3 months",date_minor_breaks="1 month",date_labels = "%m/%y")+
  labs(x="Time (month/year)",y="")
```
Overall, this analysis shows how a model with stochastic
volatility is better at describing and predicting the considered time-frame of S&P500 returns. As can be seen from the graph and perhaps unsurprisingly,
the density forecast given the stochastic volatility model represents
a particularly significant improvement in the period of extreme values of returns in March 2020.

@@

>>>>

```{r include=FALSE}
truex<-x
estimatedx<-matrix(NA,ncol=5,nrow=5)
colnames(estimatedx)<-c("N","BPF","GPF", "APF","LWF")
estimatedx[,1]<-c("20","100","1000","10000","100000")

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=4)
```

@@

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution. Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500. Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}

@@

@@

\section{Comparing Constant and Stochastic Volatility Models of Equity Returns Using the BPF}\label{Forecast comparison}

In this section, we round up the appplications chapter using the particle filter in a simple model-comparison
exercise, underlining the importance of taking stochastic
volatility into account in modeling equity returns. We follow Johannes, Korteweg and Polson (2014), comparing density forecasts through relative predictive and cumulative likelihoods.

We analyze the previously mentioned dataset of log close-to-close returns of the S&P500, from 2017 to 2021. Consider a stochastic volatility model
$$
\mathcal{M}_{SV}:\begin{array}{lc}
r_{t}=\alpha_{r}+\sigma_{r,t}\cdot v_{t}\\
\log\sigma_{r,t}^{2}=\alpha_{\sigma}+\beta_{\sigma}\cdot\log\sigma_{r,t-1}^{2}+\sigma_{\sigma}\cdot w_{t}
\end{array}\ (v_{t},w_{t})\sim\mathcal{N}_{2}(\boldsymbol{0},I_{2})
$$

Whereas the emission distribution $p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})$
is known, we approximate the predictive state distribution $p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})$
with weighted particles obtained with a Bootstrap Particle Filter.

Figure \ref{fig:cll_plot} reports the time series of the cumulative log-likelihood
given the constant volatility model $\mathcal{M}_{CV}$ relative to
the one given $\mathcal{M}_{SV}$, obtained as $\log p(y_{1:t}|\mathcal{M}_{CV})-\log p(y_{1:t}|\mathcal{M}_{SV})$.
Notice that the first difference of this series is the relative log-likelihood
of the incoming observation, $p(y_{t}|y_{1:t-1},\mathcal{M}_{CV})-p(y_{t}|y_{1:t-1},\mathcal{M}_{SV})$.

```{r cll_plot, echo=FALSE, fig.align='center', fig.cap="Relative log-likelihood of S&P500 daily log returns, CV and SV", fig.height=3, fig.pos='H', fig.width=10, results=F}
load("SV/cll_plot.Rda")
cll_plot+
  theme_bw()+
  scale_x_date(date_breaks = "3 months",date_minor_breaks="1 month",date_labels = "%m/%y")+
  labs(x="Time (month/year)",y="")
```
Overall, this analysis shows how a model with stochastic
volatility is better at describing and predicting the considered time-frame of S&P500 returns. As can be seen from the graph and perhaps unsurprisingly,
the density forecast given the stochastic volatility model represents
a particularly significant improvement in the period of extreme values of returns in March 2020.
